Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-04-23 thesis/draft.tex
some changes to thesis (draft)

\author{Submitted to\\ Dr. Jonathan Dushoff 
\\McMaster University\\Hamiltion, Ontario, Canada L8S 4K1}
\date {March 22, 2024}
\maketitle

@@

After confirming that the GCT model, which employs a fixed structure and varying parameters $a$ and $r$, achieves comparable efficacy to the LCT model, we proceeded to empirically test the model's enhanced performance. Specifically, we focused on its improved flexibility in the kappa parameter through data fitting. Initially, we utilized simulated data as a replacement for actual data. This incidence data was generated through transmission model simulations, with deviations (noise) introduced via a mean reporting ratio and a negative binomial dispersion parameter. We performed model fitting using the \textbf{bbmle} package in R, estimating the model parameters through Maximum Likelihood Estimation (MLE). Our goal was to accurately retrieve the parameter values used to generate the incidence data, assessing the fitting quality through the mean squared error (MSE).

@@

\subsection{Substage Configuration Minimally Impacts Model Fitting}
\textbf{Only temporarily. I'm going to change all of this...}

After establishing the theoretical foundation, we proceeded with model fitting to assess the GCT model's performance. The incidence data were obtained from the model simulation, and the noises were added using a mean reporting ratio of 0.9 and a negative binomial dispersion parameter of 1000. Initially, we employed the same model type for both data generation and fitting. In Figure \ref{sametypeFittingL}.A, the data, generated by the LCT model with 2 substages, were fitted using an LCT model of the identical structure, as the plot title indicates. The quality of the fitting was evaluated based on the mean squared error (MSE). With an MSE of 1.435, the fitting can be considered satisfactory. Figure \ref{sametypeFittingL}.B showcases the parameters estimated post-fitting. Conversely, employing the same incidence data but altering the fitting model's structure from an LCT with 2 substages to an LCT with 6 substages, as depicted in Figure \ref{sametypeFittingL}.C, resulted in less favorable outcomes, with the MSE escalating to 25.521. The estimated parameters, displayed in Figure \ref{sametypeFittingL}.D, showed significant discrepancies.

@@

On the other hand, employing the GCT model with 12 substages for both generating and fitting incidence data resulted in a close match, as indicated by an MSE of 0.378 (Figure \ref{sametypeFittingG}.A), and parameters post-fitting were close to the exact values (Figure \ref{sametypeFittingG}.B). In this case, altering the fitting model's structure from 12 substages to 6 substages, the model still exhibited a close match with the original in terms of both shape (Figure \ref{sametypeFittingG}.C) and parameters (Figure \ref{sametypeFittingG}.D), albeit with a slightly increased MSE (Figure \ref{sametypeFittingG}.C). This indicates that the GCT model's fitting performance is minimally impacted by changes in the model structure.

@@

\subsection{Adaptability of the GCT Model to Data Input and Structural Changes in Fitting}
Proceeding to a more challenging test, we utilized a different model type for data generation and model fitting. As demonstrated in Figure \ref{difftypeFitting}.A, the data generated by the LCT model with 2 substages were fitted using the GCT model with 12 substages. Despite the significant difference in model structure, the fitting results were reasonably favorable, with an MSE of 4.583. Altering the fitting model from a GCT with 2 substages to a GCT with 6 substages had negligible impact on the fitting outcomes (Figure \ref{difftypeFitting}.B). Similarly, changing the input data from an LCT with 2 substages to an LCT with 7 substages resulted in slightly improved fitting, though the difference was not substantial (Figure \ref{difftypeFitting}.C). These findings suggest that regardless of changes to the fitting model structure or the data used, the GCT model is capable of adapting to these variations, providing satisfactory fitting results.

@@

\subsection{Parameter Estimation}
In this section, we discuss the parameter estimation results obtained through model fitting. The true incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{2}{9}$. Random noises were added to the true incidence data using negative binomial to give us 4 different input data. The model was fitted using the GCT framework with 5 to 10 substages. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the parameter estimation is not perfect, with kappa and beta consistently overestimated, and D underestimated. However, these discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low mean squared error. It is possible for the parameter to lack confidence intervals, which could be attributed to structural inadequacies. As previously mentioned, the minimum possible value for kappa in the model with $n$ substages is $\frac{1}{n}$. The model encounters limitations when aligning with a kappa value of less than $\frac{1}{n}$. Overall, it is important to acknowledge the imprecision of parameter estimation, which will be a focus of our subsequent efforts.

\begin{figure}[h!]
    \centering
    \includegraphics[width= 1.1\textwidth]{figures/parEst.Rout.pdf}
    \caption{\textbf{GCT parameter estimation for various structures.} \\ Input data for fittings were derived from the GCT transmission model featuring 5 subtages, with $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$. Fittings were conducted for 5 to 10 substages within the GCT framework. The targeted parameter values are marked by red lines, with the 95\% confidence intervals depicted by black lines.}
    \label{parEst}
\end{figure}

@@

\jd{Try to get the habit of doing referencing in latex instead of typing section numbers (see the example I tried with sref)}
There are potential challenges that need to be addressed in the current GCT model. As discussed in Section 4.5, the parameters estimated after data fitting do not precisely reflect the actual values, yet these values are often crucial for understanding disease dynamics. For example, the basic reproduction number, $R_0$, which indicates how many secondary infections can be expected from a single infected individual in a fully susceptible population, is essential for assessing the potential for disease outbreak and epidemic spread. The calculation of $R_0$ requires the use of the contact rate $(\beta)$, recovery rate $(\gamma)$, and birth/death rate $(\mu)$. Therefore, achieving accurate parameter values from model fitting is necessary, and more study is needed.

>>>>

\author{Submitted to\\ Dr. Jonathan Dushoff 
\\McMaster University\\Hamiltion, Ontario, Canada L8S 4K1}
\date {April 23, 2024}
\maketitle

@@

After confirming that the GCT model, which employs a fixed structure and varying parameters $a$ and $r$, achieves comparable efficacy to the LCT model, we proceeded to empirically test the model's enhanced performance. Specifically, we focused on its improved flexibility in the kappa parameter through data fitting. Initially, we utilized simulated data as a replacement for actual data. This incidence data was generated through transmission model simulations, with deviations (noise) introduced via a mean reporting ratio and a negative binomial dispersion parameter. We performed model fitting using the \textbf{bbmle} package in R, estimating the model parameters through Maximum Likelihood Estimation (MLE). Our goal was to accurately retrieve the parameter values used to generate the incidence data, assessing the fitting quality through the root mean squared log error (RMSLE).

@@

\subsection{Substage Configuration Minimally Impacts Model Fitting}

After establishing the theoretical foundation, we proceeded with model fitting to assess the GCT model's performance. The incidence data were obtained from the model simulation, and the noises were added using a mean reporting ratio of 0.9 and a negative binomial dispersion parameter of 1000. Initially, we employed the same model type for both data generation and fitting. In Figure \ref{sametypeFittingL}.A, the data, generated by the LCT model with 2 substages, were fitted using an LCT model of the identical structure, as the plot title indicates. The quality of the fitting was evaluated based on the root mean squared logarithmic error (RMSLE), which measured the percentage errors between the actual and predicted value. With an RMSLE value of 1.64\%, the fitting can be considered satisfactory. Figure \ref{sametypeFittingL}.B showcases the parameters estimated post-fitting. Conversely, employing the same incidence data but altering the fitting model's structure from an LCT with 2 substages to an LCT with 6 substages, as depicted in Figure \ref{sametypeFittingL}.C, resulted in less favorable outcomes, with the RMSLE escalating to 9.30\%. The estimated parameters, displayed in Figure \ref{sametypeFittingL}.D, showed significant discrepancies.

@@

On the other hand, employing the GCT model with 12 substages for both generating and fitting incidence data resulted in a close match, as indicated by an RMSLE of 2.34\% (Figure \ref{sametypeFittingG}.A), and parameters post-fitting were close to the exact values (Figure \ref{sametypeFittingG}.B). In this case, altering the fitting model's structure from 12 substages to 6 substages, the model still exhibited a close match with the original in terms of both shape (Figure \ref{sametypeFittingG}.C) and parameters (Figure \ref{sametypeFittingG}.D). This indicates that the GCT model's fitting performance is minimally impacted by changes in the model structure.

@@

\subsection{Adaptability of the GCT Model to Data Input and Structural Changes in Fitting}
Proceeding to a more challenging test, we utilized a different model type for data generation and model fitting. As demonstrated in Figure \ref{difftypeFitting}.A, the data generated by the LCT model with 2 substages were fitted using the GCT model with 12 substages. Despite the significant difference in model structure, the fitting results were reasonably favorable, with an RMSLE of 2.44\%. Altering the fitting model from a GCT with 2 substages to a GCT with 6 substages had negligible impact on the fitting outcomes (Figure \ref{difftypeFitting}.B). Similarly, changing the input data from an LCT with 2 substages to an LCT with 7 substages resulted in slightly improved fitting, though the difference was not substantial (Figure \ref{difftypeFitting}.C). These findings suggest that regardless of changes to the fitting model structure or the data used, the GCT model is capable of adapting to these variations, providing satisfactory fitting results.

@@

\subsection{Parameter Estimation}
\label{parm est}
In this section, we discuss the parameter estimation results obtained through model fitting. The true incidence data were generated using the GCT model with 5 substages, setting the $D$ to 10 and $\kappa$ to $\frac{2}{9}$. Random noise was added to the true incidence data using a negative binomial distribution, resulting in four different input datasets (a number we specifically set for this analysis). The model was fitted using both the GCT and LCT frameworks with 5 to 10 substages using the 4 sets of data. As depicted in Figure \ref{parEst}, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the estimated parameter values derived from the LCT model, using different input data, tend to cluster together but on average vary across different substage numbers. These estimates could deviate substantially from the accurate parameter if an inappropriate structure is chosen. On the other hand, although the parameter estimation using the GCT model is not perfect—with kappa and beta being overestimated and D underestimated—the estimated values are less fluctuating across substage numbers, closer to the target values, and consistently include the target values within the 95\% confidence interval. The discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low root mean squared logarithmic error (RMSLE).

Notice that it is possible for the parameter to lack confidence intervals, which could be attributed to structural inadequacies. As previously mentioned, the minimum possible value for kappa in the model with $n$ substages is $\frac{1}{n}$. The model encounters limitations when aligning with a kappa value of less than $\frac{1}{n}$. Overall, it is important to acknowledge the imprecision of parameter estimation, which will be a focus of our subsequent efforts.

\begin{figure}[h!]
    \centering
    \includegraphics[width= 1.1\textwidth]{figures/parEst.Rout.pdf}
    \caption{\textbf{GCT parameter estimation for various structures.} \\ The four input data for fittings were derived from the GCT transmission model featuring 5 subtages, with $\beta=0.2$, $\mu=0.01$, $N=10000$, $I_0=10$. Fittings were conducted for 5 to 10 substages within the GCT and LCT frameworks. The targeted parameter values are marked by red lines, with the 95\% confidence intervals depicted by black and green lines for GCT and LCT, respectively.}
    \label{parEst}
\end{figure}

@@

There are potential challenges that need to be addressed in the current GCT model. As discussed in \sref{parm est}, the parameters estimated after data fitting do not precisely reflect the actual values, yet these values are often crucial for understanding disease dynamics. For example, the basic reproduction number, $R_0$, which indicates how many secondary infections can be expected from a single infected individual in a fully susceptible population, is essential for assessing the potential for disease outbreak and epidemic spread. The calculation of $R_0$ requires the use of the contact rate $(\beta)$, recovery rate $(\gamma)$, and birth/death rate $(\mu)$. Therefore, achieving accurate parameter values from model fitting is necessary, and more study is needed.
